---
title: "Reproducible Research: Peer Assessment 1"
author: "Victor Borges"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

```{r, Loading and preprocessing the data}
get.data <- function (){
        if (!file.exists("data")) {
        message("Creating Data folder in working directory")
        dir.create("data")
        }
        if(!file.exists("data/repdata-data-activity")) {
        fileURL <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
        download.file(fileURL, destfile = "./data/repdata-data-activity.zip")
        file.name <- "./data/repdata-data-activity.zip"
        unzip(file.name, exdir = "data")
        unlink(file.name)
        }
        else message("data already exists")
}
get.data()
activity <- read.csv("./data/activity.csv", colClasses = c("numeric", "Date", "numeric"))
activity$day <- weekdays(activity$date)
library(ggplot2)
```
# Steps Taken Per Day

### Histogram of the total number of steps taken per day

```{r, Mean total number of steps taken per day}
stepsPerDay <- aggregate(steps ~ date, activity, sum, na.rm = TRUE)
hist(stepsPerDay$steps,
     main = "Steps per day",
     xlab = "Steps",
     col = "blue",
     breaks = 7)
```

### Mean steps per day:

```{r, }
mean(stepsPerDay$steps)
```
### Median steps per day:

```{r, }
median(stepsPerDay$steps)
```
# Average Daily Activity Pattern

### Time series plot of the 5-minute interval and the average number of steps taken, averaged across all days

```{r, Average daily activity pattern}
stepsInterval <- aggregate(steps ~ interval, activity, sum, na.rm = TRUE)
plot(stepsInterval$interval,
     stepsInterval$steps, 
     type="l", xlab = "5 min - interval", 
     ylab = "Average steps", 
     main = "Average Daily Activity Pattern", 
     col = "blue")
stepsInterval$interval[which.max(stepsInterval$steps)]
```

### 5-minute interval containing the maximum number of steps (averaged across all days)

```{r, }
stepsInterval[which.max(stepsInterval$steps),]$interval
```


## Dealing with Missing Values

### Number of NA (missing) step values in the dataset

```{r, Missing values}
sum(is.na(activity))
```
### Replacing Missing Values

WE'll replace steps missing values by the mean number of steps taken for the corresponding time interval and weekday.

```{r, }
intervalDay <- aggregate(steps ~ interval + day,
                         activity,
                         mean, na.rm = TRUE)
activityInput <- merge(activity, intervalDay, 
                       by=c("interval", "day"))
activityInput <- transform(activityInput, 
                           steps.x = ifelse(is.na(steps.x),steps.y,steps.x))
activityInput <- data.frame(activityInput[,1:4])
names(activityInput) <- c("interval", "day","steps", "date")
activityInput$steps <- round(activityInput$steps, digits = 0)
activityInput <- activityInput[order(activityInput$date, activityInput$interval),]
```
### Histogram of the total number of steps taken per day with inputed values

```{r, Mean total number of steps taken per day with inputed values}
stepsPerDayInput <- aggregate(steps ~ date,
                                activityInput, 
                                sum, 
                                na.rm = TRUE)
hist(stepsPerDayInput$steps,
     main = "Steps per day",
     xlab = "Steps",
     col = "blue",
     breaks = 7)
```

### Mean steps per day with inputed values:

```{r, }
mean(stepsPerDayInput$steps)
```
### Median steps per day inputed values:

```{r, }
median(stepsPerDayInput$steps)
```
We can notice that including inputted values both median and mean total number of steps taken per day increase.

### Are there differences in activity patterns between weekdays and weekends?

```{r, }
activityInput$daytype <- ifelse(activityInput$day %in% c("Saturday", "Sunday"),"Weekend", "Weekday")
stepsPerDaytypeInput <- aggregate(steps ~ interval + daytype, activityInput, mean)
ggplot(stepsPerDaytypeInput, aes(x=interval,y=steps)) + 
geom_line(color="blue",size=0.7) + 
facet_wrap(~daytype, nrow=2, ncol=1) + 
labs(x="Interval",y="Number of Steps")
```


